---
title: ""
author: ""
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r, include=FALSE}

# Set global knit options

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE 
)

# Load libraries

library(tidyverse)
library(rvest)
library(readxl)
library(plotly)
library(flexdashboard)
library(sf)
library(shiny)
library(leaflet)
library(scales)
library(patchwork)
library(ggplot2)

```

# Exploratory Data Analysis

#### ED Visits and Hospitalizations between 2016-2018 and 2019-2021

```{r}

# Load ED data for 2016-2018

ed_2016_2018 = 
  read_csv("data/ed_2016_2018.csv")

# Load ED data for 2019-2021

ed_2019_2021 = 
  read_csv("data/ed_2019_2021.csv")

# Load NYC zipcode shapefile

nyc_zip_sf <- st_read("shapefiles/geo_export_8df5c38c-c74c-4840-9d00-da16c0201e07.shp", quiet = TRUE)

# Combine datasets

all_data <- bind_rows(
  ed_2016_2018 %>% mutate(period = "2016-2018"),
  ed_2019_2021 %>% mutate(period = "2019-2021")
)

# Create borough averages for both indicators

borough_data <- all_data %>%
  group_by(borough, period, indicator) %>%
  summarize(avg_rate = mean(zip_code_rate, na.rm = TRUE))

# Create the map_data

borough_data <- borough_data %>%
  rename(boro_name = borough)

map_data <- nyc_zip_sf %>%
  left_join(borough_data, by = "boro_name")

```

#### Emergency Department Visit Rates per 10,000 children (Aged 0-17 Years)

```{r, echo=FALSE}

# Create simple borough-level summaries

borough_summary <- all_data %>%
  group_by(borough, period, indicator) %>%
  summarize(
    avg_rate = round(mean(zip_code_rate, na.rm = TRUE), 1),
    .groups = "drop"  
  )

# Table for ED visits 

ed_summary <- borough_summary %>%
  filter(indicator == "ed_visits_rate") %>%
  pivot_wider(
    names_from = period,
    values_from = avg_rate
  ) %>%
  mutate(
    percent_change = round(((`2019-2021` - `2016-2018`) / `2016-2018` * 100), 1)
  ) %>%
  arrange(desc(`2016-2018`))

# Print ED Visits Summary

ed_summary %>%
  select(-indicator) %>%
  knitr::kable(
    col.names = c("Borough", "2016-2018", "2019-2021", "% Change")
  )

```

For ED visits, the Bronx had notably higher rates than other boroughs in both periods, with 347.7 and 180.7 per 10,000 children, respectively. All boroughs experienced substantial decreases between 2016-2018 and 2019-2021, ranging from about 43% to 48%. Despite these decreases, the Bronx's 2019-2021 rate of 180.7 remained higher than any other borough's rates from 2016-2018.


```{r}

# Create ED Visits Map 

ed_plot <- ggplot() +
  geom_sf(data = filter(map_data, indicator == "ed_visits_rate"), 
          aes(fill = avg_rate)) +
  facet_wrap(~period) +
  scale_fill_viridis_c(
    name = "Borough Average of\nED Visit Rates\nper 10,000 Children\nAged 0-17 Years", 
    option = "plasma"
  ) +
  theme_minimal() +
  labs(title = "Asthma ED Visit Rates by NYC Borough",
       subtitle = "Children aged 0-17 years") +
  theme(
    plot.title = element_text(size = 12),
    legend.position = "right",
    legend.key.height = unit(2, "cm"),
    axis.text = element_blank(),      
    axis.ticks = element_blank(),     
    panel.grid = element_blank()      
  )

# Display plot

ed_plot

```

Looking at the ED visit rates map across the two time periods (2016-2018 and 2019-2021), we can observe several key patterns:

* There appears to be a general decrease in ED visit rates from 2016-2018 to 2019-2021 across most NYC boroughs, as indicated by the darker purple colors in the later period.
* The Bronx (the northernmost borough) shows notably higher ED visit rates than other boroughs in both time periods, with rates exceeding 300 per 10,000 children aged 0-17 years in 2016-2018.
* Manhattan and Brooklyn by comparison appear to have moderate rates, while Staten Island (the southernmost borough) consistently shows lower ED visit rates in both periods.


#### Asthma hospitalization rate per 10,000 children (aged 0-17 years)

```{r, echo=FALSE}

# Table for asthma hospitalizations

hosp_summary <- borough_summary %>%
  filter(indicator == "hosp_rate") %>%
  pivot_wider(
    names_from = period,
    values_from = avg_rate
  ) %>%
  mutate(
    percent_change = round(((`2019-2021` - `2016-2018`) / `2016-2018` * 100), 1)
  ) %>%
  arrange(desc(`2016-2018`))

# Print Hospitalization Summary

hosp_summary %>%
  select(-indicator) %>%
  knitr::kable(
    col.names = c("Borough", "2016-2018", "2019-2021", "% Change")
  )

```

For hospitalizations, a similar pattern emerged compared to ED visit rates, with the Bronx having the highest rates at 58.5 and 36.8 per 10,000 children in the two periods. All boroughs showed considerable decreases between 2016-2018 and 2019-2021, ranging from approximately 32% to 40%. Queens and Staten Island consistently had the lowest hospitalization rates in both periods.

```{r}

# Create Hospitalizations Map 

hosp_plot <- ggplot() +
  geom_sf(data = filter(map_data, indicator == "hosp_rate"), 
          aes(fill = avg_rate)) +
  facet_wrap(~period) +
  scale_fill_viridis_c(
    name = "Borough Average of\nHospitalization Rates\nper 10,000 Children\nAged 0-17 Years", 
    option = "plasma"
  ) +
  theme_minimal() +
  labs(title = "Asthma Hospitalization Rates by NYC Borough",
       subtitle = "Children aged 0-17 years") +
  theme(
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),         
    axis.text = element_blank(),           
    axis.ticks = element_blank(),         
    axis.title = element_blank()         
  )

# Display plot

hosp_plot

```

Considering hospitalization rates across the two time periods (2016-2018 and 2019-2021), there are notable patterns:

* Similar to ED visits, there appears to be a general decrease in hospitalization rates from 2016-2018 to 2019-2021 across NYC boroughs, as shown by the darker purple colors in the later period.
* The Bronx again shows notably higher hospitalization rates compared to other boroughs in both periods, reaching over 50 per 10,000 children aged 0-17 years in 2016-2018.
* Geographic disparities persist in both periods, with Staten Island consistently showing lower hospitalization rates.
* When compared to the ED visit rates, the hospitalization rates are notably lower (scale of 0-50 versus 0-300), suggesting that while many children visit the ED for asthma, a smaller proportion require hospitalization.

## Fine Particulate Matter (PM2.5) 

```{r}

# Load PM data

particulate_matter = 
  read_csv("data/Air_Quality_20231208.csv")

# Produce tidied PM dataset for analysis

tidy_analysis_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "Borough") |>
  select(-unique_id, -indicator_id, -start_date) |>
    rename(borough = geo_place_name) |>
  mutate(
    period = case_when(
      time_period %in% c("Annual Average 2016", "Annual Average 2017", "Annual Average 2018") ~ "2016-2018",
      time_period %in% c("Annual Average 2019", "Annual Average 2020", "Annual Average 2021") ~ "2019-2021"
    )
  ) |>
  group_by(borough, period) |>
  summarize(
    average_pm2.5 = mean(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = period,
    values_from = average_pm2.5,
    names_prefix = "avg_pm2.5_"
  )

```

#### Average PM2.5 Levels (μg/m³) Between 2016-2018 and 2019-2021

```{r}

# Plot of average PM2.5 levels by borough between 2016-2018 and 2019-2021

tidy_analysis_pm |>
  pivot_longer(cols = -borough, names_to = "Year_Group", values_to = "PM2.5") |> 
  mutate(borough = factor(borough, levels = unique(borough[order(PM2.5)]))) |> 
  ggplot(aes(x = borough, y = PM2.5, fill = Year_Group)) +
  geom_bar(position = "dodge", stat = "identity") +
   labs(
    title = "Average PM2.5 Levels by Borough (2016-2018 vs 2019-2021)",
    x = "Borough",
    y = "PM2.5 Values",
    fill = "Time Period"
  ) 

```

**Average PM.25 levels across boroughs: 2016-2018 vs 2019-2021**

```{r}

# Print PM2.5 Summary

tidy_analysis_pm_diff <- 
  tidy_analysis_pm |>
  mutate(
    `avg_pm2.5_2016-2018` = round(`avg_pm2.5_2016-2018`, 3),
    `avg_pm2.5_2019-2021` = round(`avg_pm2.5_2019-2021`, 3),
    pm2.5_difference = round(`avg_pm2.5_2016-2018` - `avg_pm2.5_2019-2021`, 3)
  )

knitr::kable(tidy_analysis_pm_diff, col.names = c("Borough", "Average PM2.5 (2016-2018)", "Average PM2.5 (2019-2021)", "Difference in PM2.5 Levels"))

```

PM2.5 levels demonstrated consistent declines across all NYC boroughs between 2016-2018 and 2019-2021, as shown in both the bar plot and supporting data table. Manhattan maintained the highest concentrations in both periods (8.72 and 7.63 μg/m³ respectively), which is notably higher than other boroughs as evident by the tallest bars in the graph. The magnitude of improvement was remarkably uniform in the Bronx and Brooklyn, with the table showing identical reductions of 1.02 μg/m³ for both boroughs, while Queens experienced the smallest decrease at 0.72 μg/m³. Staten Island consistently maintained the lowest PM2.5 levels among all boroughs, with averages dropping from 6.68 to 5.87 μg/m³, as evidenced by the shortest bars in the plot.

#### PM2.5 Levels by United Hospital Fund (UHF42) Neighborhoods, 2016-2018 vs. 2019-2021

```{r}

# Load NYC geospatial data

geography = 
  read_excel("data/geoid_borough_name_nyc.xlsx") |>
  rename(geo_join_id = ID) |>
  mutate(geo_join_id = 
           as.character(geo_join_id))

# Tidy dataset for geospatial PM analysis

tidy_uhf42_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "UHF42") |>
  select(-unique_id, -indicator_id, -start_date) |>
  mutate(
    period = case_when(
      time_period %in% c("Annual Average 2016", "Annual Average 2017", "Annual Average 2018") ~ "2016_2018",
      time_period %in% c("Annual Average 2019", "Annual Average 2020", "Annual Average 2021") ~ "2019_2021"
    )
  ) |>
  group_by(geo_place_name, geo_join_id, period) |>
  summarize(
    average_pm2.5 = mean(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = period,
    values_from = average_pm2.5,
    names_prefix = "avg_pm2.5_"
  ) 

# Load shapefiles map for UHF data

uhf42_map <- st_read("data/UHF_42_DOHMH_2009.shp", quiet = TRUE) 

tidy_uhf42_pm <- 
  tidy_uhf42_pm |>
  rename(UHFCODE = geo_join_id) |>
  mutate(
    UHFCODE = as.double(UHFCODE))

# Combine geospatial data and PM data

map_uhf42 <- 
  uhf42_map |>
  left_join(tidy_uhf42_pm, by = "UHFCODE") |>
  pivot_longer(cols = starts_with("avg_pm2.5_"), 
               names_to = "time_period", 
               values_to = "avg_pm2.5")

# Produce heat map 

heat_map_pm2.5 <- ggplot() +
  geom_sf(data = map_uhf42, aes(fill = avg_pm2.5)) +
  facet_wrap(~time_period, ncol = 2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey50") +
  labs(
    title = "PM2.5 (ug/m3) Levels by UHF42 Neighborhood",
    fill = "PM2.5"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),
    legend.position = "right",
    legend.key.height = unit(2, "cm"),
    panel.grid = element_blank(),          
    axis.text = element_blank(),           
    axis.ticks = element_blank(),         
    axis.title = element_blank()          
  )

# Show heat map for PM2.5 

heat_map_pm2.5

```

Looking at the PM2.5 levels across the two time periods (2016-2018 and 2019-2021), we can observe a general decrease in PM2.5 concentrations across the majority of UHF42 neighborhoods, as shown by the more prevalent darker purple coloring in the 2019-2021 map indicating lower levels. Midtown Manhattan shows consistently high PM2.5 levels across both time periods, appearing as yellow and orange areas on the maps, while Staten Island maintains the lowest PM2.5 levels across all boroughs, shown by consistent dark purple coloring. 

The marked improvement in air quality from 2016-2018 to 2019-2021 likely reflects the impact of the COVID-19 pandemic, which resulted in significantly reduced traffic and industrial pollution during this period.

## Greenspace

```{r}

# Load greenspace data

greenspace_clean = read_csv("./data/Parks_Properties_20241126.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |> 
  mutate(
    borough = case_match(borough, #renamed borough according to data dictionary
      "R" ~ "Staten Island",
      "Q" ~ "Queens",
      "X" ~ "Bronx",
      "B" ~ "Brooklyn",
      "M" ~ "Manhattan"),
    typecategory = as.factor(typecategory), #converted char variables into factors
    borough = as.factor(borough)) |>
  separate(acquisitiondate, into = c("year", "month", "day"), sep = "-") |>
  select(year, borough, acres, typecategory) |> 
  filter(year < 2022 | is.na(year))

```

#### Borough and Time Period Comparison

```{r}

# Load 2016-2018 dataset

dataset1_gs = greenspace_clean |>
  filter(year < 2019 | is.na(year)) |> 
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

dataset1_gs_calc = dataset1_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2016-2018")

# Load 2019-2021 dataset

dataset2_gs = greenspace_clean |>
  filter(year < 2022 | is.na(year)) |>
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

dataset2_gs_calc = dataset2_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2019-2021")

# Combine datasets

total_acres = bind_rows(dataset1_gs_calc, dataset2_gs_calc)
  
# Total acres dataset

Greenspace_data = total_acres |>
  pivot_wider(names_from = year_group,
              values_from = acres_sum)

# Produce plot of greenspace distribution across NYC boroughs

ggplot(total_acres, aes(x = reorder(borough, acres_sum), y = acres_sum, fill = year_group)) + 
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) + 
  labs(x = "Borough", y = "Greenspace (Total Acres)", title = "Greenspaces Across Borough by Time Period", fill = "Time Period") +
  theme_minimal()

```

Analysis of greenspace distribution across NYC boroughs shows consistent patterns between 2016-2018 and 2019-2021, with notable differences in total acreage between boroughs. Staten Island maintained substantially higher greenspace coverage than other boroughs in both periods, reaching approximately 3,000 total acres in 2019-2021. Queens ranked second with around 2,200 acres, while Brooklyn and Manhattan showed moderate levels of greenspace at approximately 1,800 acres each. 

The Bronx consistently had the lowest greenspace coverage, though it showed a slight increase from around 1,100 acres in 2016-2018 to 1,200 acres in 2019-2021. The ranking of boroughs based on greenspace coverage was relatively consistent across both time periods, indicating fairly constant disparities in greenspace access over the time periods of interest.

#### Greenspace increase


```{r, echo=FALSE}

# Table of changes in greenspace

Greenspace_data |>
  mutate(
    Difference = round(`2019-2021` - `2016-2018`, 3)) |>
  arrange(desc(Difference)) |>
  knitr::kable(col.names = c("Borough", "2016-2018", "2019-2021", "Difference"))

```

Examining the changes in greenspace between 2016-2018 and 2019-2021, the data reveals variations in greenspace across boroughs and time periods. Staten Island exhibited the most significant increase, adding 258.720 acres of greenspace, followed by the Bronx with an addition of 119.625 acres. Queens showed a modest increase of 11.494 acres, while Brooklyn and Manhattan experienced minimal changes with increases of only 0.956 and 0.345 acres respectively. These differences may reflect differences in greenspace development priorities across the city's boroughs during this period.

#### Greenspace Types

```{r}

# Calculate mean acres for each type category across all years

types_nyc = greenspace_clean |>
  select(-borough) |>
  group_by(typecategory) |>
  summarise(acres_tp = mean(acres, na.rm = TRUE), .groups = "drop")

# Process data for 2016-2018 period

type1A = greenspace_clean |>
  filter(year < 2019 | is.na(year)) |> 
  group_by(year, borough, typecategory) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

# Calculate total acres by borough and type for 2016-2018

type1B = type1A |> 
  group_by(borough, typecategory) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "year1") #2016-2018

# Process data for 2019-2021 period

type2A = greenspace_clean |>
  filter(year < 2022 | is.na(year)) |>
  group_by(year, borough, typecategory) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

# Calculate total acres by borough and type for 2019-2021

type2B = type2A |> 
  group_by(borough, typecategory) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "year2") #2019-2021

# Combine both time periods and calculate total acres by type

binded = bind_rows(type1B, type2B) |>
  group_by(typecategory, year_group) |>
  summarise(acres_total = sum(acres_sum, na.rm = TRUE))

# Create bar plot comparing greenspace types between periods

binded |> 
   mutate(year_group = case_match(year_group,
      "year1" ~ "2016-2018",
      "year2" ~ "2019-2021")) |>
  ggplot(aes(x = reorder(typecategory, acres_total), y = acres_total, fill = year_group)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  labs(x = "Greenspace Types", y = "Total Acres", title = "Total Acres of NYC Greenspace by Type and Time Period", fill = "Time Period") +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

Analysis of greenspace types across NYC reveals flagship parks as the dominant category, representing the largest total acreage in both 2016-2018 and 2019-2021 periods. Community parks and nature areas represent the second and third largest greenspace types respectively in both time periods.
The distribution of greenspace types remained largely consistent between the two time periods, with minimal changes in most categories. 

Smaller greenspace types such as gardens, plazas, and historic house parks account for relatively minimal acreage in comparison to the larger park categories. Most categories showed slight increases or remained stable across the two time periods, which suggests efforts towards the preservation and/or expansion of various greenspace types throughout the city.

```{r, echo=FALSE}

# Create table showing change in acres between time periods

Total_acres_table_test = binded |>
 pivot_wider(names_from = year_group,
             values_from = acres_total) |>
 mutate(across(where(is.numeric), ~round(., 3)),
        diff = round(year2 - year1, 3))

# Filter for only positive changes and sort by largest increase 

table_type = Total_acres_table_test |>
 filter(diff > 0) |>
 arrange(desc(diff))

# Produce table with greenspace types between years

knitr::kable(table_type, col.names = c("Greenspace Type", "2016-2018", "2019-2021", "Difference"))

```

Between 2016-2018 and 2019-2021, neighborhood parks showed the largest expansion with an increase of 259.617 acres, followed by managed sites which grew by 119.095 acres. Nature areas experienced a notable increase of 32.860 acres, while undeveloped spaces, waterfront facilities, gardens, and playgrounds showed minimal changes, with increases ranging from 0.299 to 1.186 acres. The remaining 14 greenspace types maintained their original acreage with no changes observed during this period. This pattern indicates that recent greenspace development has primarily focused on expanding neighborhood parks and managed sites while preserving existing spaces in other categories.

**Top 3 Greenspace Types (Total Acres) Across Boroughs**

```{r, echo=FALSE, message = FALSE}

# Create initial dataset of acres by type, borough, and year group

perBor = bind_rows(type1B, type2B) |>
  group_by(typecategory, year_group, borough) |>
  summarise(acres_total = sum(acres_sum, na.rm = TRUE)) 

# Process data for each borough separately
# For each:
# 1. Filter for specific borough
# 2. Round acreage to 2 decimal places
# 3. Reshape data to wide format with years as columns
# 4. Sort by year1 values in descending order
# 5. Filter to keep only the largest spaces (threshold varies by borough)

Bronx_perBor = perBor |> 
  filter(borough == "Bronx") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 200) 

Brook_perBor = perBor |> 
  filter(borough == "Brooklyn") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 500) 

Man_perBor = perBor |> 
  filter(borough == "Manhattan") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 200) 

Q_perBor = perBor |> 
  filter(borough == "Queens") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 900) 

SI_perBor = perBor |> 
  filter(borough == "Staten Island") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 760)

# Combine all borough data and format for final table

top3_bor_gstype = bind_rows(Bronx_perBor, Brook_perBor, Q_perBor, Man_perBor, SI_perBor) |>
  select(borough, typecategory, year1, year2) |>
  rename(acres = year1) |>
  select(-year2) |>
  group_by(borough) |>
  mutate(rank = rank(-acres)) |>
  ungroup() |>
  mutate(
    rank = case_match(rank,
      1 ~ "First",
      2 ~ "Second",
      3 ~ "Third"),
    typecategory = paste(typecategory, " (", acres, ")", sep = "")) |>
    select(-acres) |>
  pivot_wider(
    names_from = rank,
    values_from = typecategory) 

# Create formatted table with clear column names

knitr::kable(top3_bor_gstype, col.names = c("Borough", "Highest", "Second Highest", "Third Highest"))
 
```

The table shows the top three greenspace types by total acreage across New York City's five boroughs, revealing distinct patterns in how each borough maintains its green spaces. Staten Island has the largest single greenspace with a Nature Area of 1690.97 acres, showing a focus on preserving natural environments. Manhattan maintains relatively smaller spaces, with its largest being a Community Park at 1098.39 acres. The outer boroughs feature major destination parks called Flagship Parks - which NYC Parks targets for long-term planning and community input. 

The Bronx and Brooklyn both incorporate significant Parkway acreage as part of New York State's historic system of landscaped roads. Queens stands out with the largest Flagship Park (1631.46 acres), while Brooklyn shows diversity in its green spaces with both large Community Parks and Parkways.

## Temperature

```{r}

# Load and clean raw temperature data for each borough

manhattan = read_csv("data/manhattan.csv",skip = 3) |>
  mutate(Borough = "Manhattan") |> 
  janitor::clean_names()

bronx = read_csv("data/bronx.csv",skip = 3) |>
  mutate(Borough = "Bronx") |> 
  janitor::clean_names()

brooklyn = read_csv("data/brooklyn.csv",skip = 3) |>
  mutate(Borough = "Brooklyn") |> 
  janitor::clean_names()

queens = read_csv("data/queens.csv",skip = 3) |>
  mutate(Borough = "Queens") |> 
  janitor::clean_names()

staten_island = read_csv("data/staten_island.csv",skip = 3) |>
  mutate(Borough = "Staten Island") |> 
  janitor::clean_names()

# Combine temperature data from all boroughs and format dates/times

all_temp = bind_rows(manhattan, bronx, brooklyn, queens, staten_island) |> 
  mutate(borough = as.factor(borough)) |> 
  mutate(date = as.Date(time, format = "%d/%m/%Y")) |> 
  select(time:temperature_2m_mean, borough, date) |> 
  mutate(years = year(date)) |> 
  mutate(months = as.Date(paste0(format(date, "%Y-%m"), "-01"), format = "%Y-%m-%d"))

# Calculate average temperatures by borough and year group

all_temp_borough = all_temp |> 
  mutate(date = as.Date(time, format = "%d/%m/%Y")) |> 
  mutate(data_years = ifelse(year(date) %in% 2016:2018, "2016-2018", "2019-2021")) |> 
  select(temperature_2m_max:data_years) |> 
  group_by(data_years, borough) |> 
  summarise(avg_temperature = mean(temperature_2m_mean, na.rm = TRUE))

```

#### Monthly Mean temperature by Borough

```{r}

# Produce plot of monthly Mean temperature

all_temp_month = all_temp |> 
  group_by(months, borough) |> 
  summarise(avg_month = mean(temperature_2m_mean, na.rm = TRUE))

ggplot(all_temp_month, aes(x = months, y = avg_month, color = borough)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Monthly Mean Temperature (2016-2021) by Borough",
    x = "Month",
    y = "Average Temperature (Celsius)",
    color = "Borough"
  ) +
  theme_minimal()

```

The plot shows monthly mean temperatures across New York City's five boroughs from 2016 to 2021. There is a clear seasonal pattern, with temperatures peaking around 25°C in summer months and dropping to near 0°C in winter months. Notably, the temperature patterns are remarkably similar across all boroughs, with the lines closely overlapping throughout the time period. This suggests that despite their different geographic locations and urban landscapes, the boroughs experience very similar temperature patterns throughout the year. The consistency of these patterns persists across all years shown in the dataset.

#### Max, Min, Mean temperature trend by year and borough

This 5 plots shows the highest temperature mean, lowest temperature mean, and total mean temperature by year and borough.

```{r}

# Mean temperature data

mean_temp_year = all_temp |> 
  group_by(years, borough) |> 
  summarize(
    high_temp = mean(temperature_2m_max, na.rm = TRUE),
    low_temp = mean(temperature_2m_min, na.rm = TRUE),
    mean_temp = mean(temperature_2m_mean, na.rm = TRUE)
  ) |> 
  arrange(years, borough) 

# Pivot to long format

long_data <- mean_temp_year |> 
  pivot_longer(
    cols = c(high_temp, low_temp, mean_temp),
    names_to = "Temperature_Type",            
    values_to = "Temperature"             
  )

# Produce plot of max, Min, Mean temperature trend by year and borough

ggplot(long_data, aes(x = years, y = Temperature, color = Temperature_Type, group = Temperature_Type)) +
  geom_line(size = 1) +                   
  geom_point(size = 2) +              
  labs(
    title = "Mean Temperature Trends by Year and Borough",
    x = "Year",
    y = "Temperature (°C)",       
    color = "Mean Temperature Type"
  ) +
  scale_color_manual(
    values = c("high_temp" = "red", "low_temp" = "blue", "mean_temp" = "darkgreen"),
    labels = c("high_temp" = "Highest Temperature", 
               "low_temp" = "Lowest Temperature", 
               "mean_temp" = "Mean Temperature")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ borough, scales = "free_x") +
  theme(aspect.ratio = 0.6)
  
```

This plot shows the temperature trends across New York City's five boroughs from 2016 to 2021, tracking the highest, lowest, and mean temperatures for each borough. Each borough shows similar patterns, with highest temperatures consistently around 17°C, mean temperatures around 12.5°C, and lowest temperatures around 8°C. The temperature patterns remain relatively stable across the six-year period, with only minor fluctuations in all temperature types. 

All boroughs show a slight upward trend in temperatures from 2019 to 2020, particularly noticeable in the highest temperature lines. Despite their different locations and urban characteristics, the boroughs appear to have very similar temperature profiles throughout the study period, suggesting consistent climate patterns across New York City.

#### Average temperature change by borough (2016-2021)

```{r}

# Produce average temperature table with rounded values

mean_temp_year |> 
  select(years, borough, mean_temp) |> 
  mutate(mean_temp = round(mean_temp, 3)) |>
  pivot_wider(
    names_from = borough,
    values_from = mean_temp
  ) |>
  knitr::kable(col.names = c("Years", "Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"))

```

The table presents annual mean temperatures across New York City's five boroughs from 2016 to 2021. There is a notable downward trend from 2016 to 2019, with temperatures decreasing by approximately 0.5-0.7°C across all boroughs. In 2020, there was a marked increase in temperatures, with all boroughs reaching their highest values since 2016, averaging around 12.8°C. 

Temperatures remained elevated in 2021, though slightly lower than 2020 peaks. Throughout the period, Brooklyn and Staten Island consistently maintained slightly higher average temperatures compared to the other boroughs, while Manhattan generally recorded the lowest average temperatures.

#### Average temperature by time period and borough


```{r}

# Plot for average temperature by year group and borough

all_temp_borough |>
  rename(Borough = borough) |>
ggplot(aes(x = data_years, y = avg_temperature, color = Borough)) +
  geom_point(size = 3) +
  geom_line(aes(group = Borough), size = 1) +
  labs(title = "Average Temperature by Time Period and Borough",
       x = "Time period",
       y = "Average Temperature") +
  theme_minimal()

```

This graph illustrates the change in average temperatures across New York City's boroughs between two time periods: 2016-2018 and 2019-2021. All boroughs experienced an increase in average temperature between these periods, with Manhattan and Bronx showing the most pronounced increases. Brooklyn consistently maintained the highest average temperatures across both periods, starting at approximately 12.42°C and rising to 12.62°C, while Manhattan, despite its significant increase, recorded the lowest temperatures in both periods. The parallel upward trends across all boroughs suggest a consistent citywide warming pattern, with temperatures rising by roughly 0.2-0.3°C between the two time periods.

## Heat Vulnerability Index (HVI)

```{r}
#import merged file from data preparation page
full_data2 = read_csv("data/NYC_total.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
    mutate(
    data_years = as.factor(data_years),
    borough = as.factor(borough),
    indicator = as.factor(indicator),
    zip_code = as.factor(zip_code))

#convert categorical variables into factors

# Calculate average HVI scores for each borough
# This helps understand overall heat vulnerability patterns across boroughs

daa = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |>
  group_by(borough) |> 
  summarise(hvi_mean = mean(hvi))

# Produce table
knitr::kable(daa, digits = 3, col.names = c("Borough", "Mean HVI"))
```


```{r dataset, include = FALSE}

# Data ED Visits (2016-2018)

full_data2_ed_1 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "ed_visits_rate") |> 
  filter(data_years == "2016-2018")

## Datasets for Graphing (datasets 1 & 2):
## Changing HVI to numeric

# Data ED Visits (2019-2021)

full_data2_ed_2 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "ed_visits_rate") |> 
  filter(data_years == "2019-2021")

# Data Hospitalizations (2016-2018)

full_data2_hosp_1 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "hosp_rate") |>  
  filter(data_years == "2016-2018")

# Data Hospitalizations (2019-2021)

full_data2_hosp_2 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "hosp_rate") |>  
  filter(data_years == "2019-2021")

## Datasets for Statistical Analysis (datasets 3 & 4):
## Keeping HVI as a factor variable


# Data ED Visits (2016-2018)

full_data2_ed_3 = full_data2 |> 
  filter(indicator == "ed_visits_rate") |>  # filtering only ed_visits
  filter(data_years == "2016-2018")

# Data ED Visits (2019-2021)

full_data2_ed_4 = full_data2 |> 
  filter(indicator == "ed_visits_rate") |>  # filtering only ed_visits
  filter(data_years == "2019-2021")

# Data Hospitalizations (2016-2018)

full_data2_hosp_3 = full_data2 |> 
  filter(indicator == "hosp_rate") |>  #filtering only hospitlization rate
  filter(data_years == "2016-2018")

# Data Hospitalizations (2019-2021)

full_data2_hosp_4 = full_data2 |> 
  filter(indicator == "hosp_rate") |>  #filtering only hospitlization rate
  filter(data_years == "2019-2021")

```

The mean Heat Vulnerability Index varies considerably across NYC boroughs, with the Bronx showing the highest average (4.07), followed by Brooklyn (3.65), Queens (3.27), Manhattan (2.42), and Staten Island (2.07). This suggests that heat vulnerability is not evenly distributed across the city, with some boroughs experiencing systematically higher heat vulnerability than others. The notable variation in HVI means across boroughs indicates that geographic and demographic factors play an important role in heat vulnerability distribution in NYC.

```{r}

# Plot for mean HVI scores by borough

borough_plot = ggplot(daa, aes(x = reorder(borough, -hvi_mean), y = hvi_mean)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(hvi_mean, 2)), 
            vjust = -0.5, 
            size = 3.5) +
  labs(title = "Mean Heat Vulnerability Index (HVI) by Borough",
       x = "Borough",
       y = "Mean HVI Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  ylim(0, max(daa$hvi_mean) * 1.1)  # Add some space above bars for labels

# Display the plot

borough_plot

```

The above plot provides a visual representation of these HVI disparities, with a noticeable difference between the Bronx and Staten Island - a gap of 2 points on the HVI scale.