---
title: ""
author: ""
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Predictor-Outcome Associations
The plots below show the associations between each predictor and the outcome of interest (ED Visits Rate and Hospitalization Rate).

```{r libraries, include=FALSE}
library(tidyverse)
library(patchwork)
```

```{r merged file, include=FALSE}
#import merged file from data preparation page
full_data2 = read_csv("data/NYC_total.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
    mutate(
    data_years = as.factor(data_years),
    borough = as.factor(borough),
    indicator = as.factor(indicator),
    zip_code = as.factor(zip_code))

#convert categorical variables into factors
```

## PM2.5  

```{r pm2.5 file, include=FALSE}
#clean merged file for pm2.5 and outcome comparisons across boroughs and year groups
assoc_pm2_5 = full_data2 |>
  rename(Year = data_years, Borough = borough) |> #renamed it to reflect on plots
  select(Year, Borough, indicator, pm2_5_level, borough_rate)
```

#### Year Group Comparison

```{r pm2.5 ed, echo=FALSE, message=FALSE}
#clean data for Borough level ED visit rates and pm2.5
ED_assoc_pm2_5 = assoc_pm2_5 |>
  filter(indicator == "ed_visits_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and obtain values of interest
EDmodels_pm2_5 = ED_assoc_pm2_5 |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ pm2_5_level, data = .)) |> #linear model
  mutate( 
    slope = coef(model)[2], #calc slope, intercept, R², and correlation coefficient (R)
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = sign(slope) * sqrt(r_squared)) |>
   select(-model, -intercept)

# Plot the data and regression lines
EDplot_pm2_5 = ggplot(ED_assoc_pm2_5, aes(x = pm2_5_level, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "PM2.5 Levels and ED Visits Rate", 
       x = "PM2.5 Level (ug/m3)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

#plot
EDplot_pm2_5

#Make table of results
knitr::kable(EDmodels_pm2_5, digits = 3)
```

Comparing between the two year groups, there are lower ED visist rates and lower PM2.5 levels in the 2019-2021 year group. 

The correlation coefficient for both year groups are similar. Data from 2016 to 2018 indicates a positive association between PM2.5 levels and ED visit rate (R = 0.377). Data from 2019 to 2021 also shows a positive association between these two variables (R = 0.323). Thus, decreased PM2.5 levels is weakly associated with decreased hospitalization rate for both year groups.

```{r pm2.5 hosp, echo=FALSE, message=FALSE}
#clean data for Borough level hospitalization rates and pm2.5
hosp_assoc_pm2_5 = assoc_pm2_5 |>
  filter(indicator == "hosp_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and obtain values of interest
hospmodels_pm2_5 = hosp_assoc_pm2_5 |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ pm2_5_level, data = .)) |>
  mutate(
    slope = coef(model)[2],
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = sign(slope) * sqrt(r_squared)) |>
   select(-model, -intercept)

# Plot the data and regression lines
hosplot_pm2_5 = ggplot(hosp_assoc_pm2_5, aes(x = pm2_5_level, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "PM2.5 Levels and Hospitalization Rate", 
       x = "PM2.5 Levels (ug/m3)", 
       y = "Hospitalization Rate") +
  theme(legend.position = "top")

#plot 
hosplot_pm2_5

#Make table of results
knitr::kable(hospmodels_pm2_5, digits = 3)

```

Similar to the trend comparing PM2.5 levels and ED visit rates, hospitalization rates and PM2.5 levels are also lower in the 2019-2021 year group.

The correlation coefficient (r) for both year groups is 0.414, which indicates a moderate relationship between the two variables for the year groups. Decreased PM2.5 levels is thus moderately associated with decreased hospitalization rate for both year groups.

#### Borough and Year Comparison

```{r pm2.5 boroughyear, echo=FALSE, message=FALSE}

GS_ED_pm2_5 = assoc_pm2_5 |>
  filter(indicator == "ed_visits_rate") |> #look at ed visit rate
  ggplot(aes(x = pm2_5_level, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "PM2.5 Levels(ug/m3)", y = "ED Visit Rate", 
       title = "NYC PM2.5 Levels and ED Visit Rate") +
  theme_minimal()

GS_ED_pm2_5
```

For all boroughs, PM2.5 levels and ED visit rates are higher in 2016 to 2018 than in 2019 to 2021. Compared to the other boroughs, decrease in ED visit rate per PM2.5 level is most dramatic in the Bronx.

```{r pm2.5 HR, echo=FALSE, message=FALSE}

GS_HR_pm2_5 = assoc_pm2_5 |>
  filter(indicator == "hosp_rate") |> #look at hospitalization rate
  ggplot(aes(x = pm2_5_level, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "PM2.5 Levels(ug/m3)", y = "Hospitalization Rate", 
       title = "NYC PM2.5 Levels and Hospitalization Rate") +
  theme_minimal()

GS_HR_pm2_5

```

Similar to the trend seen between PM2.5 levels and ED visit rates, PM2.5 levels and hospitalization rates are higher in 2016 to 2018 than in 2019 to 2021. Bronx also experienced the most dramatic decrease in hospitalization rate per PM2.5 level. 

## Greenspace

```{r GS file, include=FALSE}
#clean merged file for greenspace and outcome comparisons across boroughs and year groups
GS_assoc = full_data2 |>
  rename(Year = data_years, Borough = borough) |> #renamed it to reflect on plots
  select(Year, Borough, indicator, acres_sum, borough_rate)
```

#### Year Group Comparison

```{r ed, echo=FALSE, message=FALSE}
#clean data for Borough level ED visit rates and greenspace
ED_assoc = GS_assoc |>
  filter(indicator == "ed_visits_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and obtain values of interest
#calc slope, intercept, R², and correlation coefficient (R)
EDmodels = ED_assoc |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ acres_sum, data = .)) |> 
  mutate( 
    slope = coef(model)[2], 
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = sign(slope) * sqrt(r_squared)) |>
   select(-model, -intercept)

# Plot the data and regression lines
EDplot = ggplot(ED_assoc, aes(x = acres_sum, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Greenspace and ED Visits Rate", 
       x = "Greenspace (Total Acres)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

#plot
EDplot

#Make table of results
knitr::kable(EDmodels, digits = 3)
```

The correlation coefficient for both year groups are similar. Data from 2016 to 2018 indicates a negative association between greenspace and ED visit rate (r = -0.905). Data from 2019 to 2021 also shows a negative association between these two variables (r = -0.830). Thus, increased greenspace is strongly associated with decreased ED visit rate for both year groups.

```{r hosp, echo=FALSE, message=FALSE}

#clean data for Borough level hospitalization rates and greenspace
hosp_assoc = GS_assoc |>
  filter(indicator == "hosp_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and extract results
hospmodels = hosp_assoc |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ acres_sum, data = .)) |>
  mutate(
    slope = coef(model)[2],
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = sign(slope) * sqrt(r_squared)) |>
   select(-model, -intercept)

# Plot the data and regression lines
hosplot = ggplot(hosp_assoc, aes(x = acres_sum, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Greenspace and Hospitalization Rate", 
       x = "Greenspace (Total Acres)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

#plot 
hosplot

#Make table of results
knitr::kable(hospmodels, digits = 3)

```

Similar to the trend comparing greenspace and ED visit rates, the correlation coefficient for both year groups are similar. Data from 2016 to 2018 indicates a negative association between greenspace and hospitalization rate (r = -0.861). Data from 2019 to 2021 also shows a negative association between these two variables (r = -0.793). Thus, **increased greenspace is strongly associated with decreased hospitalization rate for both year groups.

#### Borough and Year Comparison

```{r, echo=FALSE, message=FALSE}

GS_ED = GS_assoc |>
  filter(indicator == "ed_visits_rate") |> #look at ed visit rate
  ggplot(aes(x = acres_sum, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Greenspace (Total Acres)", y = "ED Visit Rate", 
       title = "NYC Greenspace and ED Visit Rate") +
  theme_minimal()

GS_ED
```

In Bronx, there is a dramatic decrease in ED visit rate with an increase in total acres of greenspace between the two year groups. This decrease is also seen for Staten Island. The association is less obvious for Brooklyn, Manhattan, and Queens. 


```{r, echo=FALSE, message=FALSE}

GS_HR = GS_assoc |>
  filter(indicator == "hosp_rate") |> #look at hospitalization rate
  ggplot(aes(x = acres_sum, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Greenspace (Total Acres)", y = "Hospitalization Rate", 
       title = "NYC Greenspace and Hospitalization Rate") +
  theme_minimal()

GS_HR

```

When comparing between the two year groups, the association for borough-level greenspace and ED visit rate is similar to trend for borough-level greenspace and hospitalization rate.

## Temperature

```{r echo = FALSE, message = FALSE}
temp_association = full_data2 |> 
  select(data_years, avg_temperature, indicator, borough, borough_rate) |> 
  distinct(data_years, avg_temperature, indicator, borough, borough_rate) |>
  rename(Year = data_years, Borough = borough)

temp_association_ed = temp_association |> 
  filter(indicator == "ed_visits_rate") 

temp_association_hosp = temp_association |> 
  filter(indicator == "hosp_rate")
```

#### Year Group Comparison

```{r echo = FALSE, message = FALSE}
tempmodel = temp_association_ed |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ avg_temperature, data = .)) |> #linear model
  mutate( 
    slope = coef(model)[2], #calc slope, intercept, R², and correlation coefficient (R)
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = -sqrt(r_squared)) |> #negative correlation
   select(-model, -intercept)

# Plot the data and regression lines
tempplot = ggplot(temp_association_ed, aes(x = avg_temperature, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Temperature and ED Visits Rate", 
       x = "Average Temperature (°C)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

tempplot
knitr::kable(tempmodel, digits = 3)
```

The correlation coefficient for the year groups are different. Data from 2016 to 2018 indicates a weak negative association between temperature and ED visit rate (r = -0.323). The correlation coefficient for 2019 to 2021 (r = -0.039) indicates no association between these two variables. 

We can infer that temperature is not a significant factor for ED visits.

```{r echo = FALSE, message = FALSE}
tempmodel2 = temp_association_hosp |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ avg_temperature, data = .)) |> #linear model
  mutate( 
    slope = coef(model)[2], #calc slope, intercept, R², and correlation coefficient (R)
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = -sqrt(r_squared)) |> #negative correlation
   select(-model, -intercept)

# Plot the data and regression lines
tempplot2 = ggplot(temp_association_hosp, aes(x = avg_temperature, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Temperature and Hospitalization Rate", 
       x = "Average Temperature (°C)", 
       y = "Hospitalization Rate") +
  theme(legend.position = "top")

tempplot2
knitr::kable(tempmodel2, digits = 3)
```

Data from 2016 to 2018 indicates a weak negative association between temperature and hospitalization rate (r = -0.348). The correlation coefficient for 2019 to 2021 (r = -0.120) indicates a very week negative association between these two variables. 

We can infer that temperature is not a significant factor for Hospitalization.

#### Borough and Year Comparison

```{r echo=FALSE, message=FALSE}
ggplot(temp_association, aes(x = avg_temperature, y = borough_rate, color = Borough)) +
  geom_point(size = 3) +
  facet_wrap(~indicator) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship between Temperature and Rates by Borough",
    x = "Average Temperature (°C)",
    y = "Rate"
  ) +
  theme_minimal()
```

The plot shows that the higher temperature is associated with lower rate of ED visit and hospitalization.

## Heat Vulnerability Index (HVI)

```{r dataset, include = FALSE}

# Data ED Visits (2016-2018)

full_data2_ed_1 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "ed_visits_rate") |> 
  filter(data_years == "2016-2018")

## Datasets for Graphing (datasets 1 & 2):
## Changing HVI to numeric

# Data ED Visits (2019-2021)

full_data2_ed_2 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "ed_visits_rate") |> 
  filter(data_years == "2019-2021")

# Data Hospitalizations (2016-2018)

full_data2_hosp_1 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "hosp_rate") |>  
  filter(data_years == "2016-2018")

# Data Hospitalizations (2019-2021)

full_data2_hosp_2 = full_data2 |> 
  mutate(hvi = as.numeric(hvi)) |> 
  filter(indicator == "hosp_rate") |>  
  filter(data_years == "2019-2021")

## Datasets for Statistical Analysis (datasets 3 & 4):
## Keeping HVI as a factor variable


# Data ED Visits (2016-2018)

full_data2_ed_3 = full_data2 |> 
  filter(indicator == "ed_visits_rate") |>  # filtering only ed_visits
  filter(data_years == "2016-2018")

# Data ED Visits (2019-2021)

full_data2_ed_4 = full_data2 |> 
  filter(indicator == "ed_visits_rate") |>  # filtering only ed_visits
  filter(data_years == "2019-2021")

# Data Hospitalizations (2016-2018)

full_data2_hosp_3 = full_data2 |> 
  filter(indicator == "hosp_rate") |>  #filtering only hospitlization rate
  filter(data_years == "2016-2018")

# Data Hospitalizations (2019-2021)

full_data2_hosp_4 = full_data2 |> 
  filter(indicator == "hosp_rate") |>  #filtering only hospitlization rate
  filter(data_years == "2019-2021")

```

```{r, fig.height=10, fig.width=12, echo=FALSE, message=FALSE}

# Create visualization plots showing relationship between HVI scores and health outcomes
# Four plots examine ED visits and hospitalizations across both time periods

# Create plots

plot1 = ggplot(full_data2_ed_1, aes(x = hvi, y = zip_code_rate)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "ED Visit rate per HVI score (2016-2018)", 
       x = "HVI", y = "ED Visit Rate") +
  theme_minimal()

plot2 = ggplot(full_data2_ed_2, aes(x = hvi, y = zip_code_rate)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "ED Visit rate per HVI score (2019-2021)", 
       x = "HVI", y = "ED Visit Rate") +
  theme_minimal()

plot3 = ggplot(full_data2_hosp_1, aes(x = hvi, y = zip_code_rate)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Hospitalization rate per HVI score (2016-2018)", 
       x = "HVI", y = "Hospitalization Rate") +
  theme_minimal()

plot4 = ggplot(full_data2_hosp_2, aes(x = hvi, y = zip_code_rate)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Hospitalization rate per HVI score (2019-2021)", 
       x = "HVI", y = "Hospitalization Rate") +
  theme_minimal()

# Combine plots with adjusted height and width
(plot1 + plot2) / (plot3 + plot4) +
  patchwork::plot_layout(heights = c(3, 3), widths = c(3, 3))

```

The scatter plots show consistent positive relationships between Heat Vulnerability Index and both pediatric asthma health outcomes across time periods. ED visit rates show a stronger association with HVI than hospitalization rates, as observed in steeper trend lines. The relationship appears more pronounced in the 2016-2018 period, with ED visits ranging from 0 to approximately 700 per 10,000 across HVI scores, compared to 0-350 in 2019-2021. 

While the overall patterns persist in 2019-2021, the rates are lower, particularly for ED visits. The considerable vertical spread in the data points at each HVI level suggests that other factors beyond heat vulnerability influence pediatric asthma health outcomes.