---
title: ""
author: ""
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Predictor-Outcome Associations

```{r libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
```

```{r merged file, include=FALSE}

#import merged file from data preparation page
full_data2 = read_csv("data/NYC_total.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
    mutate(
    data_years = as.factor(data_years),
    borough = as.factor(borough),
    indicator = as.factor(indicator),
    zip_code = as.factor(zip_code))

#convert categorical variables into factors
```
## PM2.5  

### ED Admissions

* Odds Ratios for ED admissions, by time period

```{r}


```

### ED Hospitalizations

* Odds Ratios for ED hospitalizations, by time period

```{r}



```

## Greenspace
### ED Admissions


```{r GS file, include=FALSE}
#clean merged file for greenspace and outcome comparisons across boroughs and year groups
GS_assoc = full_data2 |>
  rename(Year = data_years, Borough = borough) |> #renamed it to reflect on plots
  select(Year, Borough, indicator, acres_sum, borough_rate)
```

#### Year Group Comparison

```{r ed, echo=FALSE, message=FALSE}
#clean data for Borough level ED visit rates and greenspace
ED_assoc = GS_assoc |>
  filter(indicator == "ed_visits_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and obtain values of interest
EDmodels = ED_assoc |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ acres_sum, data = .)) |> #linear model
  mutate( 
    slope = coef(model)[2], #calc slope, intercept, R², and correlation coefficient (R)
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = -sqrt(r_squared)) |> #negative correlation
   select(-model, -intercept)

# Plot the data and regression lines
EDplot = ggplot(ED_assoc, aes(x = acres_sum, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Greenspace and ED Visits Rate", 
       x = "Greenspace (Total Acres)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

#plot
EDplot

#Make table of results
knitr::kable(EDmodels, digits = 3)
```

For both year groups, total acres of borough-level greenspace is negatively associated with borough-level ED visit rate. Increased greenspace is associated with decreased hospitalization rate. 

```{r hosp, echo=FALSE, message=FALSE}

#clean data for Borough level hospitalization rates and greenspace
hosp_assoc = GS_assoc |>
  filter(indicator == "hosp_rate") |>
  distinct() |>
  select(-indicator)

# Fit linear models for each year group and extract results
hospmodels = hosp_assoc |>
  group_by(Year) |>
  do(model = lm(borough_rate ~ acres_sum, data = .)) |>
  mutate(
    slope = coef(model)[2],
    intercept = coef(model)[1],
    r_squared = summary(model)$r.squared,
    r = -sqrt(r_squared)) |> #negative correlation
   select(-model, -intercept)

# Plot the data and regression lines
hosplot = ggplot(hosp_assoc, aes(x = acres_sum, y = borough_rate, group = Year)) + 
  geom_point(aes(color = Year)) + 
  geom_smooth(method = "lm", aes(color = Year, group = Year), se = FALSE, linetype = "solid") +
  theme_minimal() +
  labs(title = "Greenspace and Hospitalization Rate", 
       x = "Greenspace (Total Acres)", 
       y = "ED Visits Rate") +
  theme(legend.position = "top")

#plot 
hosplot

#Make table of results
knitr::kable(hospmodels, digits = 3)

```


For both year groups, total acres of borough-level greenspace is negatively associated with borough-level hospitalization rate. Increase in greenspace is associated with decrease hospitalization rate. 


#### Borough and Year Comparison

```{r, echo=FALSE, message=FALSE}

GS_ED = GS_assoc |>
  filter(indicator == "ed_visits_rate") |> #look at ed visit rate
  ggplot(aes(x = acres_sum, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Greenspace (Total Acres)", y = "ED Visit Rate", 
       title = "NYC Greenspace and ED Visit Rate") +
  theme_minimal()

GS_ED
```

In Bronx, an increase in total acres of greenspace between the two year groups is negatively associated with ED visit rate. This is also seen for Staten Island. The association is less obvious for Brooklyn, Manhattan, and Queens. 


```{r, echo=FALSE, message=FALSE}

GS_HR = GS_assoc |>
  filter(indicator == "hosp_rate") |> #look at hospitalization rate
  ggplot(aes(x = acres_sum, y = borough_rate, color = Borough)) + 
  geom_point(aes(shape = Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Greenspace (Total Acres)", y = "Hospitalization Rate", 
       title = "NYC Greenspace and Hospitalization Rate") +
  theme_minimal()

GS_HR

```

When comparing between the two year groups, the association for borough greenspace and ED Visit Rate is similar to trend for borough greenspace and hospitalization rate.


### ED Hospitalizations

* Odds Ratios for ED hospitalizations, by time period

```{r}



```

## Temperature

### ED Admissions

* Odds Ratios for ED admissions, by time period

```{r}



```

### ED Hospitalizations

* Odds Ratios for ED hospitalizations, by time period

```{r}



```
