---
title: "Report"
author: ""
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: hide
knitr::opts_chunk$set:
  message: FALSE
  warning: FALSE
---

# Motivation

Children’s health and development is impacted by the surrounding environment, particularly in dense urban environments like New York City (NYC). Poor air quality due to urban air pollution can influence and have long-term consequences for child and adolescent respiratory health. The challenges of urban environmental health are further complicated by climate change, particularly through the urban heat island effect, which can increase the risk of heat-related illnesses (Heaviside et al., 2017). Limited research shows that elevated temperatures due to the urban heat island effect are associated with increased emergency department (ED) visits among children (Sheffield et al., 2018). 

These health outcomes may be influenced by the availability of green spaces, which can help filter air pollutants and provide cooler areas for shelter from heat. Importantly, environmental health risks disproportionately affect children from low-income neighborhoods and communities of color, where there is often limited access to green spaces and higher exposure to environmental stressors (Corburn et al., 2006).

Through this project, we aim to investigate how these interconnected environmental determinants- air quality, green space availability, and urban heat patterns - vary across NYC boroughs and their potential associations with children’s health, specifically focusing on ED asthma admissions and hospitalizations. We believe that understanding these relationships is crucial for informing urban planning policies and public health initiatives aimed at improving children’s health in New York City.

# Initial Questions

#### 1. How do pediatric asthma outcomes and environmental factors vary across NYC?

* How do asthma ED admission and hospitalization rates vary across NYC boroughs between 2016-2018 and 2019-2021?
* What is the distribution of environmental factors (PM2.5, temperature, greenspace) across NYC during these periods?

#### 2. What relationships emerge between environmental factors and pediatric asthma rates across NYC?

* Do areas with higher ED admission and hospitalization rates align with particular environmental patterns?
* Are there geographic clusters where both environmental inequities and higher asthma rates persist across time periods?
* What are the estimated associations between environmental factors (PM2.5, temperature, greenspace) and pediatric asthma ED rates, accounting for borough-level and temporal variation?

# Data

```{r setup, include=FALSE}

# Set global knit options

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

```

```{r}

# Load libraries 

library(tidyverse)
library(sf)
library(ggplot2)
library(readxl)

```

# Exploratory Analysis

Our exploratory analysis examined relevant environmental and health variables for pediatric asthma across New York City's five boroughs between 2016-2018 and 2019-2021. We explored two health indicators: pediatric asthma emergency department (ED) visits and pediatric asthma hospitalizations. We further examined three environmental factors of interest to our research aims, specifically distributions of fine particulate matter (PM2.5), greenspace, and temperature patterns across New York City.

## Asthma-Related Healthcare Utilization
### Emergency Department (ED) Visits

```{r}

# Load ED data for 2016-2018

ed_2016_2018 = 
  read_csv("data/ed_2016_2018.csv")

# Load ED data for 2019-2021

ed_2019_2021 = 
  read_csv("data/ed_2019_2021.csv")

# Combine datasets

all_data <- bind_rows(
  ed_2016_2018 %>% mutate(period = "2016-2018"),
  ed_2019_2021 %>% mutate(period = "2019-2021")
)

# Create borough averages for both indicators

borough_data <- all_data %>%
  group_by(borough, period, indicator) %>%
  summarize(avg_rate = mean(zip_code_rate, na.rm = TRUE))

# Create simple borough-level summaries

borough_summary <- all_data %>%
  group_by(borough, period, indicator) %>%
  summarize(
    avg_rate = round(mean(zip_code_rate, na.rm = TRUE), 1),
    .groups = "drop"  
  )

# Table for ED visits 

ed_summary <- borough_summary %>%
  filter(indicator == "ed_visits_rate") %>%
  pivot_wider(
    names_from = period,
    values_from = avg_rate
  ) %>%
  mutate(
    percent_change = round(((`2019-2021` - `2016-2018`) / `2016-2018` * 100), 1)
  ) %>%
  arrange(desc(`2016-2018`))

# Print ED Visits Summary

ed_summary %>%
  select(-indicator) %>%
  knitr::kable(
    col.names = c("Borough", "2016-2018", "2019-2021", "% Change")
  )

```

**Table 1. Borough Average Asthma ED Visit Rates per 10,000 Children (aged 0-17 Years)**

```{r}

# Load NYC zipcode shapefile

nyc_zip_sf <- st_read("shapefiles/geo_export_8df5c38c-c74c-4840-9d00-da16c0201e07.shp", quiet = TRUE)

# Create the map_data

borough_data <- borough_data %>%
  rename(boro_name = borough)

map_data <- nyc_zip_sf %>%
  left_join(borough_data, by = "boro_name")

# Create ED Visits Map 

ed_plot <- ggplot() +
  geom_sf(data = filter(map_data, indicator == "ed_visits_rate"), 
          aes(fill = avg_rate)) +
  facet_wrap(~period) +
  scale_fill_viridis_c(
    option = "plasma"
  ) +
  theme_minimal() +
  labs(title = "Pediatric Asthma ED Visit Rates by NYC Borough") +
  theme(
    plot.title = element_text(size = 12),
    legend.position = "right",
    legend.key.height = unit(2, "cm"),
    axis.text = element_blank(),      
    axis.ticks = element_blank(),     
    panel.grid = element_blank()      
  )

# Display plot

ed_plot

```

**Figure 1. Borough Average ED Visit Rates per 10,000 Children (aged 0-17 Years)**

Analysis of ED visit rates showed significant disparities dependent on borough and time period. The Bronx, the northernmost borough of NYC, consistently showed the highest rates across both periods, with 347.7 and 180.7 visits per 10,000 children respectively. All boroughs experienced substantial decreases between 2016-2018 and 2019-2021, with reductions ranging from 43% to 48%. 

Despite this overall decline, the Bronx's 2019-2021 rate remained higher than any other borough's 2016-2018 rates. Geographic visualization highlighted persistent health disparities between boroughs, with the Bronx showing consistently higher visit rates compared to other areas of the city, including Manhattan to its south, Queens to the east, and the more distant southern boroughs of Brooklyn and Staten Island.

The most striking change between the two periods is the apparent reduction in ED visits during 2019-2021, which could potentially be related to the COVID-19 pandemic's impact on healthcare utilization patterns.

### Hospitalizations

```{r, echo=FALSE}

# Table for asthma hospitalizations

hosp_summary <- borough_summary %>%
  filter(indicator == "hosp_rate") %>%
  pivot_wider(
    names_from = period,
    values_from = avg_rate
  ) %>%
  mutate(
    percent_change = round(((`2019-2021` - `2016-2018`) / `2016-2018` * 100), 1)
  ) %>%
  arrange(desc(`2016-2018`))

# Print Hospitalization Summary

hosp_summary %>%
  select(-indicator) %>%
  knitr::kable(
    col.names = c("Borough", "2016-2018", "2019-2021", "% Change")
  )

```

**Table 2. Borough Average Asthma Hospitalizations per 10,000 Children (aged 0-17 Years)**

Hospitalization patterns largely reflected ED visit trends. The Bronx again led with the highest rates at 58.5 and 36.8 per 10,000 children across the two periods. All boroughs showed considerable decreases between periods, ranging from 32% to 40%. Queens and Staten Island had the lowest hospitalization rates throughout both time periods. 

Notably, the ratio between ED visits and hospitalizations (approximately 6:1) suggests that while many children visit the ED for asthma, a smaller proportion require hospitalization.


```{r}

# Create Hospitalizations Map 

hosp_plot <- ggplot() +
  geom_sf(data = filter(map_data, indicator == "hosp_rate"), 
          aes(fill = avg_rate)) +
  facet_wrap(~period) +
  scale_fill_viridis_c(
    option = "plasma"
  ) +
  theme_minimal() +
  labs(title = "Pediatric Asthma Hospitalization Rates by NYC Borough") +
  theme(
    plot.title = element_text(size = 12),
    panel.grid = element_blank(),         
    axis.text = element_blank(),           
    axis.ticks = element_blank(),         
    axis.title = element_blank()         
  )

# Display plot

hosp_plot

```

**Figure 2. Borough Average Asthma Hospitalization Rates per 10,000 Children (aged 0-17 Years)**

Figures 1 and 2 reveal persistent borough-level disparities in both ED visits and hospitalizations, with higher rates of healthcare utilization in the Bronx reflecting an unequal distribution of child asthma burden in New York City. These patterns suggest underlying socioeconomic and environmental factors affecting child asthma rates across different boroughs, with communities from lower-resourced parts of New York City bearing a disproportionate share of the health burden.


## Environmental Determinants
### Air Quality Patterns (PM2.5)


```{r}

# Load PM data

particulate_matter = 
  read_csv("data/Air_Quality_20231208.csv")

# Produce tidied PM dataset for analysis

tidy_analysis_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "Borough") |>
  select(-unique_id, -indicator_id, -start_date) |>
    rename(borough = geo_place_name) |>
  mutate(
    period = case_when(
      time_period %in% c("Annual Average 2016", "Annual Average 2017", "Annual Average 2018") ~ "2016-2018",
      time_period %in% c("Annual Average 2019", "Annual Average 2020", "Annual Average 2021") ~ "2019-2021"
    )
  ) |>
  group_by(borough, period) |>
  summarize(
    average_pm2.5 = mean(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = period,
    values_from = average_pm2.5,
    names_prefix = "avg_pm2.5_"
  )

```

```{r}

# Plot of average PM2.5 levels by borough between 2016-2018 and 2019-2021

tidy_analysis_pm |>
  pivot_longer(cols = -borough, names_to = "Year_Group", values_to = "PM2.5") |> 
  mutate(borough = factor(borough, levels = unique(borough[order(PM2.5)]))) |> 
  ggplot(aes(x = borough, y = PM2.5, fill = Year_Group)) +
  geom_bar(position = "dodge", stat = "identity") +
   labs(
    title = "Average PM2.5 Levels by Borough (2016-2018 vs 2019-2021)",
    x = "Borough",
    y = "PM2.5 Values",
    fill = "Year Group"
  ) 

```

**Figure 3. Average PM.25 levels across boroughs: 2016-2018 vs 2019-2021**

```{r}

# Print PM2.5 Summary

tidy_analysis_pm_diff <- 
  tidy_analysis_pm |>
  mutate(
    `avg_pm2.5_2016-2018` = round(`avg_pm2.5_2016-2018`, 3),
    `avg_pm2.5_2019-2021` = round(`avg_pm2.5_2019-2021`, 3),
    pm2.5_difference = round(`avg_pm2.5_2016-2018` - `avg_pm2.5_2019-2021`, 3)
  )

knitr::kable(tidy_analysis_pm_diff, col.names = c("Borough", "Average PM2.5 (2016-2018)", "Average PM2.5 (2019-2021)", "Difference in PM2.5 Levels"))

```

**Table 3. Borough Average ED Visit Rates per 10,000 Children (aged 0-17 Years)**

Fine particulate matter (PM2.5) concentrations show distinct spatial patterns that partially align with asthma outcomes. Manhattan, despite not having the highest asthma rates, consistently recorded the highest PM2.5 concentrations (8.72 μg/m³ in 2016-2018, decreasing to 7.63 μg/m³ in 2019-2021). This interesting divergence from reported asthma ED visit and hospitalization rates suggests that while air quality is a significant factor, other environmental and social determinants influence pediatric asthma outcomes in New York City.


```{r}

# Load NYC geospatial data

geography = 
  read_excel("data/geoid_borough_name_nyc.xlsx") |>
  rename(geo_join_id = ID) |>
  mutate(geo_join_id = 
           as.character(geo_join_id))

# Tidy dataset for geospatial PM analysis

tidy_uhf42_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "UHF42") |>
  select(-unique_id, -indicator_id, -start_date) |>
  mutate(
    period = case_when(
      time_period %in% c("Annual Average 2016", "Annual Average 2017", "Annual Average 2018") ~ "2016_2018",
      time_period %in% c("Annual Average 2019", "Annual Average 2020", "Annual Average 2021") ~ "2019_2021"
    )
  ) |>
  group_by(geo_place_name, geo_join_id, period) |>
  summarize(
    average_pm2.5 = mean(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = period,
    values_from = average_pm2.5,
    names_prefix = "avg_pm2.5_"
  ) 

# Load shapefiles map for UHF data

uhf42_map <- st_read("data/UHF_42_DOHMH_2009.shp", quiet = TRUE) 

tidy_uhf42_pm <- 
  tidy_uhf42_pm |>
  rename(UHFCODE = geo_join_id) |>
  mutate(
    UHFCODE = as.double(UHFCODE))

# Combine geospatial data and PM data

map_uhf42 <- 
  uhf42_map |>
  left_join(tidy_uhf42_pm, by = "UHFCODE") |>
  pivot_longer(cols = starts_with("avg_pm2.5_"), 
               names_to = "time_period", 
               values_to = "avg_pm2.5")

# Produce heat map 

heat_map_pm2.5 <- ggplot() +
  geom_sf(data = map_uhf42, aes(fill = avg_pm2.5)) +
  facet_wrap(~time_period, ncol = 2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey50") +
  labs(
    title = "PM2.5 (ug/m3) Levels by UHF42 Neighborhood",
    fill = "PM2.5"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),
    legend.position = "right",
    legend.key.height = unit(2, "cm"),
    panel.grid = element_blank(),          
    axis.text = element_blank(),           
    axis.ticks = element_blank(),         
    axis.title = element_blank()          
  )

# Show heat map for PM2.5 

heat_map_pm2.5

```

**Figure 4. PM2.5 levels by United Hospital Fund (UHF) Neighborhoods**

The uniform reduction in PM2.5 levels across boroughs between the two periods (particularly the identical 1.02 μg/m³ reduction in both the Bronx and Brooklyn) suggests city-wide factors, possibly including urban policy interventions or COVID-19-related activity reductions, affected air quality improvements. Based on UHF42 neighborhood-level analysis, Midtown Manhattan has consistently high PM2.5 levels across both time periods. The PM2.5 hotspots in Midtown Manhattan suggest urban factors, such as traffic patterns, industrial activities and building density, likely play a key role in determining local air quality.

### Greenspace

```{r}

# Load greenspace data

greenspace_clean = read_csv("./data/Parks_Properties_20241126.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |> 
  mutate(
    borough = case_match(borough, #renamed borough according to data dictionary
      "R" ~ "Staten Island",
      "Q" ~ "Queens",
      "X" ~ "Bronx",
      "B" ~ "Brooklyn",
      "M" ~ "Manhattan"),
    typecategory = as.factor(typecategory), #converted char variables into factors
    borough = as.factor(borough)) |>
  separate(acquisitiondate, into = c("year", "month", "day"), sep = "-") |>
  select(year, borough, acres, typecategory) |> 
  filter(year < 2022 | is.na(year))

```


```{r}

# Load 2016-2018 dataset

dataset1_gs = greenspace_clean |>
  filter(year < 2019 | is.na(year)) |> 
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

dataset1_gs_calc = dataset1_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2016-2018")

# Load 2019-2021 dataset

dataset2_gs = greenspace_clean |>
  filter(year < 2022 | is.na(year)) |>
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

dataset2_gs_calc = dataset2_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2019-2021")

# Combine datasets

total_acres = bind_rows(dataset1_gs_calc, dataset2_gs_calc)
  
# Total acres dataset

Greenspace_data = total_acres |>
  pivot_wider(names_from = year_group,
              values_from = acres_sum)

# Produce plot of greenspace distribution across NYC boroughs

ggplot(total_acres, aes(x = reorder(borough, acres_sum), y = acres_sum, fill = year_group)) + 
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) + 
  labs(x = "Borough", y = "Greenspace (Total Acres)", title = "Greenspaces Across Borough by Year Group") +
  theme_minimal()

```

**Figure 5. Greenspace distribution across boroughs: 2016-2018 vs 2019-2021**

```{r, echo=FALSE}

# Table of changes in greenspace

Greenspace_data |>
  mutate(
    Difference = round(`2019-2021` - `2016-2018`, 3)) |>
  arrange(desc(Difference)) |>
  knitr::kable(col.names = c("Borough", "2016-2018", "2019-2021", "Difference"))

```

**Table 4. Difference in Greenspace distribution across boroughs**

Our analysis highlights substantial inequities in greenspace access across New York City. Staten Island's dominance in both total coverage (approximately 3,000 acres) and increased greenspace (258.72 acres) stands in contrast to the Bronx, which showed the lowest coverage despite a modest increase of 119.625 acres. This apparent inverse relationship between greenspace availability and asthma rates - with the Bronx having the highest asthma ED visit and hospitalization rates and lowest greenspace coverage - suggests a potential environmental health relationship warranting further investigation.


```{r}

# Calculate mean acres for each type category across all years

types_nyc = greenspace_clean |>
  select(-borough) |>
  group_by(typecategory) |>
  summarise(acres_tp = mean(acres, na.rm = TRUE), .groups = "drop")

# Process data for 2016-2018 period

type1A = greenspace_clean |>
  filter(year < 2019 | is.na(year)) |> 
  group_by(year, borough, typecategory) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

# Calculate total acres by borough and type for 2016-2018

type1B = type1A |> 
  group_by(borough, typecategory) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "year1") #2016-2018

# Process data for 2019-2021 period

type2A = greenspace_clean |>
  filter(year < 2022 | is.na(year)) |>
  group_by(year, borough, typecategory) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

# Calculate total acres by borough and type for 2019-2021

type2B = type2A |> 
  group_by(borough, typecategory) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "year2") #2019-2021

# Combine both time periods and calculate total acres by type

binded = bind_rows(type1B, type2B) |>
  group_by(typecategory, year_group) |>
  summarise(acres_total = sum(acres_sum, na.rm = TRUE))

# Create bar plot comparing greenspace types between periods

binded |> 
   mutate(year_group = case_match(year_group,
      "year1" ~ "2016-2018",
      "year2" ~ "2019-2021")) |>
  ggplot(aes(x = reorder(typecategory, acres_total), y = acres_total, fill = year_group)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  labs(x = "Greenspace Types", y = "Total Acres", title = "Total Acres of NYC Greenspace by Type and Year Group") +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

**Figure 6. Greenspace types across boroughs: 2016-2018 vs 2019-2021**

```{r, echo=FALSE}

# Create table showing change in acres between time periods

Total_acres_table_test = binded |>
 pivot_wider(names_from = year_group,
             values_from = acres_total) |>
 mutate(across(where(is.numeric), ~round(., 3)),
        diff = round(year2 - year1, 3))

# Filter for only positive changes and sort by largest increase 

table_type = Total_acres_table_test |>
 filter(diff > 0) |>
 arrange(desc(diff))

# Produce table with greenspace types between years

knitr::kable(table_type, col.names = c("Greenspace Type", "2016-2018", "2019-2021", "Difference"))

```

The analysis of greenspace types suggests varying priorities in green developments across New York City. The substantial expansion of neighborhood parks (259.617 acres) and managed sites (119.095 acres) indicates focused efforts to increase accessible green spaces. However, the minimal changes in smaller categories like gardens and plazas suggests that large-scale greenspace development may not adequately address neighborhood-level access needs, which could be more meaningful for boroughs like the Bronx.

**Table 5. Top 3 Greenspace Types (Total Acres) Across Boroughs**

```{r, echo=FALSE, message = FALSE}

# Create initial dataset of acres by type, borough, and year group

perBor = bind_rows(type1B, type2B) |>
  group_by(typecategory, year_group, borough) |>
  summarise(acres_total = sum(acres_sum, na.rm = TRUE)) 

# Process data for each borough separately
# For each:
# 1. Filter for specific borough
# 2. Round acreage to 2 decimal places
# 3. Reshape data to wide format with years as columns
# 4. Sort by year1 values in descending order
# 5. Filter to keep only the largest spaces (threshold varies by borough)

Bronx_perBor = perBor |> 
  filter(borough == "Bronx") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 200) 

Brook_perBor = perBor |> 
  filter(borough == "Brooklyn") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 500) 

Man_perBor = perBor |> 
  filter(borough == "Manhattan") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 200) 

Q_perBor = perBor |> 
  filter(borough == "Queens") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 900) 

SI_perBor = perBor |> 
  filter(borough == "Staten Island") |>
  mutate(acres_total = round(acres_total, 2)) |>
  pivot_wider(
    names_from = year_group,
    values_from = acres_total) |>
  arrange(desc(year1)) |>
  filter(year1 > 760)

# Combine all borough data and format for final table

top3_bor_gstype = bind_rows(Bronx_perBor, Brook_perBor, Q_perBor, Man_perBor, SI_perBor) |>
  select(borough, typecategory, year1, year2) |>
  rename(acres = year1) |>
  select(-year2) |>
  group_by(borough) |>
  mutate(rank = rank(-acres)) |>
  ungroup() |>
  mutate(
    rank = case_match(rank,
      1 ~ "First",
      2 ~ "Second",
      3 ~ "Third"),
    typecategory = paste(typecategory, " (", acres, ")", sep = "")) |>
    select(-acres) |>
  pivot_wider(
    names_from = rank,
    values_from = typecategory) 

# Create formatted table with clear column names

knitr::kable(top3_bor_gstype, col.names = c("Borough", "Highest", "Second Highest", "Third Highest"))
 
```

While the Bronx contains major Flagship Parks and Parkways among its largest green spaces, these types of spaces may not provide the same respiratory health benefits as smaller, more accessible neighborhood parks distributed throughout residential areas where children live and play. Despite the expansion of neighborhood parks (259.617 acres) and managed sites (119.095 acres) across the city, the concentration of greenspace in large, potentially less accessible formats like Flagship Parks may limit their protective effects against childhood asthma. Staten Island's lower asthma rates and large Nature Area (1690.97 acres) present an interesting contrast, though this association may reflect broader factors such as lower population density or better air quality rather than the specific type of greenspace.

### Temperature

```{r}

# Load and clean raw temperature data for each borough

manhattan = read_csv("data/manhattan.csv",skip = 3) |>
  mutate(Borough = "Manhattan") |> 
  janitor::clean_names()

bronx = read_csv("data/bronx.csv",skip = 3) |>
  mutate(Borough = "Bronx") |> 
  janitor::clean_names()

brooklyn = read_csv("data/brooklyn.csv",skip = 3) |>
  mutate(Borough = "Brooklyn") |> 
  janitor::clean_names()

queens = read_csv("data/queens.csv",skip = 3) |>
  mutate(Borough = "Queens") |> 
  janitor::clean_names()

staten_island = read_csv("data/staten_island.csv",skip = 3) |>
  mutate(Borough = "Staten Island") |> 
  janitor::clean_names()

# Combine temperature data from all boroughs and format dates/times

all_temp = bind_rows(manhattan, bronx, brooklyn, queens, staten_island) |> 
  mutate(borough = as.factor(borough)) |> 
  mutate(date = as.Date(time, format = "%d/%m/%Y")) |> 
  select(time:temperature_2m_mean, borough, date) |> 
  mutate(years = year(date)) |> 
  mutate(months = as.Date(paste0(format(date, "%Y-%m"), "-01"), format = "%Y-%m-%d"))

# Calculate average temperatures by borough and year group

all_temp_borough = all_temp |> 
  mutate(date = as.Date(time, format = "%d/%m/%Y")) |> 
  mutate(data_years = ifelse(year(date) %in% 2016:2018, "2016-2018", "2019-2021")) |> 
  select(temperature_2m_max:data_years) |> 
  group_by(data_years, borough) |> 
  summarise(avg_temperature = mean(temperature_2m_mean, na.rm = TRUE))

```


```{r}

# Produce plot of monthly Mean temperature

all_temp_month = all_temp |> 
  group_by(months, borough) |> 
  summarise(avg_month = mean(temperature_2m_mean, na.rm = TRUE))

ggplot(all_temp_month, aes(x = months, y = avg_month, color = borough)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Monthly Mean Temperature (2016-2021) by Borough",
    x = "Month",
    y = "Average Temperature (Celsius)",
    color = "Borough"
  ) +
  theme_minimal()

```

Temperature analysis shows differences across New York City's boroughs between 2016 and 2021. While seasonal variations show consistent patterns across boroughs (peaks around 25°C and lows near 0°C), the slight but persistent temperature differences between boroughs suggest varying urban heat island effects. 


```{r}

# Mean temperature data

mean_temp_year = all_temp |> 
  group_by(years, borough) |> 
  summarize(
    high_temp = mean(temperature_2m_max, na.rm = TRUE),
    low_temp = mean(temperature_2m_min, na.rm = TRUE),
    mean_temp = mean(temperature_2m_mean, na.rm = TRUE)
  ) |> 
  arrange(years, borough) 

# Produce average temperature table with rounded values

mean_temp_year |> 
  select(years, borough, mean_temp) |> 
  mutate(mean_temp = round(mean_temp, 3)) |>
  pivot_wider(
    names_from = borough,
    values_from = mean_temp
  ) |>
  knitr::kable(col.names = c("Years", "Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"))

```

Brooklyn and Staten Island's slightly consistently higher average temperatures, compared to Manhattan's lower averages, may reflect differences in building density and greenspace coverage.

The observed warming trend between periods (0.2-0.3°C increase) aligns with broader climate change patterns but may also reflect the effects of localized urban developments. This warming pattern, combined with limited greenspace in asthma hotspot areas like the Bronx, suggests potential intersecting environmental health challenges.



#### Heat Vulnerability Index

While our exploratory analysis examined temperature patterns across NYC boroughs, our final modeling approach ultimately employed the Heat Vulnerability Index (HVI) rather than raw temperature measurements. This decision was driven by several key considerations. First, our exploratory analysis revealed that temperature patterns were very similar across all boroughs, with only slight variations in average temperatures (0.2-0.3°C differences). This uniformity in temperature measurements meant it would not be easy to explain the substantial geographic variations we observed in asthma outcomes. Second, HVI would be able to represent neighborhood-level variations in heat vulnerability that raw temperature data alone would unlikely show based on the data obtained.

Instead, we opted to use the Heat Vulnerability Index, which provides a more comprehensive measure of the impact of heat stress on different communities. The HVI incorporates not only environmental factors but also crucial socio-demographic characteristics that influence a community's resilience to heat stress. These include language vulnerability (relevant for NYC's diverse neighborhoods), socioeconomic vulnerability, environmental and urban vulnerability, and elderly isolation - factors that may indirectly affect childhood asthma outcomes through the social determinants of health, including housing quality, access to cooling infrastructure, and overall community resources.


# Correlations

# Regression Analysis

# References

Corburn, J., Osleeb, J., & Porter, M. (2006). Urban asthma and the neighbourhood environment in New York City. Health & place, 12(2), 167-179.

Heaviside, C., Macintyre, H., & Vardoulakis, S. (2017). The urban heat island: implications for health in a changing environment. Current environmental health reports, 4, 296-305.

Sheffield, P. E., Herrera, M. T., Kinnee, E. J., & Clougherty, J. E. (2018). Not so little differences: variation in hot weather risk to young children in New York City. Public health, 161, 119–126. https://doi.org/10.1016/j.puhe.2018.06.004


